---
title: "Time Series of dLST"
author: "Jonathan Lanz"
date: "2026-05-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE, 
                      fig.path = here::here("fig/"), 
                      dev = "png", 
                      dpi = 300
)
```

```{r, include=FALSE}
# load libraries and functions
library(ggplot2)
library(here)
library(scico)
library(lubridate)
library(cowplot)
library(dplyr)
source(here::here("R", "get_lst_plot.R"))
source(here::here("R", "get_stats_label.R"))# is executed inside get_lst_plot

# load data
model_data_clean <- readRDS(here::here("data", "model_data_clean.rds")) |>
  filter(!land_cover_type %in% c(11,17))
spatial_predictions_lst <- readRDS(here::here("data", "spatial_predictions_lst.rds"))
dry_predictions_lst <- readRDS(here::here("data", "dry_predictions_lst.rds"))
wet_days <- readRDS(here::here("data", "wet_days.rds"))
ndvi_jjas_181920_ch <- readRDS(here::here("data", "ndvi_jjas_181920_ch.rds"))

dry_predictions_lst_ndvi <- dry_predictions_lst |>
  left_join(
    ndvi_jjas_181920_ch |> mutate(lat = round(lat, 4), lon = round(lon, 4)),
    by = c("lat", "lon", "date")
  )
```

## Timeline of dLST and PCWD on driest pixel 

Let's first look at the dLST and PCWD on the driest pixel, which is at the following coordinates:

```{r dlst_timeline_driest_pixel, message=FALSE, warning=FALSE, echo=FALSE}
# get driest pixel
driest_pixel_data <- spatial_predictions_lst |>
  left_join(
    model_data_clean |> dplyr::select(lat, lon, date, pcwd_mm),
    by = c("lat", "lon", "date")
  ) |>
  group_by(lat, lon) |>
  mutate(mean_pcwd = mean(pcwd_mm, na.rm = TRUE)) |>
  ungroup() |>
  dplyr::filter(lat == lat[which.max(mean_pcwd)],
                lon == lon[which.max(mean_pcwd)])

#print driest pixel
driest_pixel_data |>
  dplyr::distinct(lat, lon, mean_pcwd)

#plot
driest_pixel_data |>
  ggplot(aes(x = date)) +
  geom_line(aes(y = lst_delta_smooth), colour = "firebrick") +
  geom_smooth(aes(y = lst_delta_smooth), method = "lm", colour = "darkred", 
            linetype = "dotted", linewidth = 0.5, se = FALSE) +
  geom_line(aes(y = pcwd_mm / 20), colour = "steelblue", linetype = "dashed") +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  scale_y_continuous(
    name     = "ΔLST [K]",
    sec.axis = sec_axis(~ . * 20, name = "PCWD [mm]")
  ) +
  facet_wrap(~lubridate::year(date), ncol = 1, scales = "free_x") +
  labs(title = "ΔLST and PCWD at driest pixel by PCWD",
       x = "Date") +
  theme_minimal()
```

From looking at this plot, it seems that dLST is not a good water stress indicator. However, this result might also indicate that the driest pixel by PCWD is not equal to the pixel where plants have highest water stress, meaning that PCWD is actually a worse water stress indicator than dLST.

## Timeline of mean dLST and PCWD across entire study area

```{r dlst_timeline_all, message=FALSE, warning=FALSE, echo=FALSE}
# plot timeline of dLST and PCWD
spatial_predictions_lst |>
  left_join(model_data_clean |> dplyr::select(lat, lon, date, pcwd_mm),
            by = c("lat", "lon", "date")) |>
  group_by(date) |>
  summarise(
    mean_delta = mean(lst_delta_smooth, na.rm = TRUE),
    mean_pcwd  = mean(pcwd_mm, na.rm = TRUE)
  ) |>
  ggplot(aes(x = date)) +
  geom_line(aes(y = mean_delta), colour = "firebrick") +
  geom_smooth(aes(y = mean_delta), method = "lm", colour = "firebrick", 
            linetype = "dotted", linewidth = 0.5, se = FALSE) +
  geom_line(aes(y = mean_pcwd / 20), colour = "steelblue", linetype = "dashed") +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  scale_y_continuous(
    name     = "Mean ΔLST [K]",
    sec.axis = sec_axis(~ . * 20, name = "Mean PCWD [mm]")
  ) +
  facet_wrap(~lubridate::year(date), ncol = 1, scales = "free_x") +
  labs(title = "Mean ΔLST and PCWD across study area",
       x = "Date") +
  theme_minimal()

```
This graphs shows quite well that LST reacts faster to condition changes than PCWD. The advantage of (P)CWD is that it shows the cumulation and thus the severity of drought better. LST has a higher variance and, although there is a rising trend, fails to indicate the cumulation aspect. 

## Timeline of mean dLST and NDVI across entire study area

```{r dlst_timeline_ndvi, message=FALSE, warning=FALSE, echo=FALSE}
spatial_predictions_lst |>
  mutate(lat = round(lat, 4), lon = round(lon, 4)) |>
  left_join(
    ndvi_jjas_181920_ch |> mutate(lat = round(lat, 4), lon = round(lon, 4)),
    by = c("lat", "lon", "date")
  ) |>
  group_by(date) |>
  summarise(
    mean_delta = mean(lst_delta_smooth, na.rm = TRUE),
    mean_ndvi  = mean(ndvi, na.rm = TRUE)
  ) |>
  ggplot(aes(x = date)) +
  geom_line(aes(y = mean_delta), colour = "firebrick") +
  geom_line(aes(y = (mean_ndvi - 0.5) * 20), colour = "darkgreen", linetype = "dashed") +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  scale_y_continuous(
    name     = "Mean ΔLST [K]",
    sec.axis = sec_axis(~ . / 20 + 0.5, name = "Mean NDVI")
  ) +
  facet_wrap(~lubridate::year(date), ncol = 1, scales = "free_x") +
  labs(title = "Mean ΔLST and NDVI across study area",
       x = "Date") +
  theme_minimal()
```

