---
title: "Model validation"
author: "Jonathan Lanz"
date: "2026-05-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE, 
                      fig.path = here::here("fig/"), 
                      dev = "png", 
                      dpi = 300
)
```


```{r}
#load libraries
library(here)
library(ggplot2)
library(dplyr)
library(tidyr)
library(tidymodels)
library(ranger)


#load model
rf_fit <- readRDS(here::here("data", "rf_fit.rds"))

#load data
validation_data_clean <- readRDS(here::here("data", "validation_data_clean.rds"))

ndvi_jjas_181920_ch <- readRDS(here::here("data", "ndvi_jjas_181920_ch.rds"))
dry_predictions_lst <- readRDS(here::here("data", "dry_predictions_lst.rds"))

#add ndvi to dry predictions
dry_predictions_lst_ndvi <- dry_predictions_lst |>
  left_join(
    ndvi_jjas_181920_ch |> mutate(lat = round(lat, 4), lon = round(lon, 4)),
    by = c("lat", "lon", "date")
  )

#prepare validation data
pcwd_threshold <- 55

val_wet_days <- validation_data_clean |>
  group_by(date) |>
  summarise(mean_pcwd = mean(pcwd_mm, na.rm = TRUE)) |>
  filter(mean_pcwd <= pcwd_threshold) |>
  pull(date)

val_data <- validation_data_clean |>
  filter(date %in% val_wet_days) |>
  dplyr::select(lst_kelvin, elevation, land_cover_type, tot_ssrd, mean_t2m, rin, mean_d2m) |>
  mutate(land_cover_type = factor(land_cover_type)) |>
  drop_na() |>
  filter(!land_cover_type %in% c(11,17))
```

## Temporal cross-validation of the random forest model

```{r}
# evaluate
set.seed(42) # for reproducibility
predictions1 <- rf_fit |>
  predict(val_data) |>
  bind_cols(val_data)

#evaluate without LC type "Urban and Built-Up Land"
val_data_no_urban <- val_data |>
  filter(land_cover_type != 13)

set.seed(42) # for reproducibility
predictions2 <- rf_fit |>
  predict(val_data_no_urban) |>
  bind_cols(val_data_no_urban)

metrics(predictions1, truth = lst_kelvin, estimate = .pred)
metrics(predictions2, truth = lst_kelvin, estimate = .pred)

predictions1 |>
  summarise(bias = mean(.pred - lst_kelvin, na.rm = TRUE))

predictions2 |>
  summarise(bias = mean(.pred - lst_kelvin, na.rm = TRUE))
```

Repeated validation runs result in an R^2 of 0.65 to 0.7, which is solid. Let's see which predictor is the most important:

```{r}
set.seed(42)
rf_fit |>
  extract_fit_parsnip() |>
  vip::vip()
```

Air temperature is by far the most important predictor. This is not optimal, since air temperature is the only one of the predictors that is reversely influenced by LST. However, it was a result that could be expected. The really small influence of TRI is likely due to the low resolution, which doesn't capture much topographic variance. It is expected that the influence of TRI will rise significantly with higher resolution.


```{r}
predictions1 |>
  dplyr::mutate(residual = lst_kelvin - .pred) |>
  ggplot(aes(x = .pred, y = residual)) +
  geom_point(alpha = 0.1, size = 0.3, colour = "grey40") +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "firebrick") +
  labs(title = "Residuals vs Predicted LST",
       x = "Predicted LST [K]", y = "Residual [K]") +
  theme_minimal()

# distribution of residuals
predictions2 |>
  dplyr::mutate(residual = lst_kelvin - .pred) |>
  ggplot(aes(x = residual)) +
  geom_histogram(bins = 50, fill = "steelblue", alpha = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "firebrick") +
  labs(title = "Distribution of residuals",
       x = "Residual [K]", y = "n") +
  theme_minimal()
```

The model seems to have negligible bias. Residuals are centered around 0 and are distributed along a Gaussian bell curve. 

## Physical comparison with NDVI

```{r}
df_plot <- dry_predictions_lst_ndvi |>
  tidyr::drop_na(ndvi, lst_delta)

dens <- MASS::kde2d(df_plot$ndvi, df_plot$lst_delta, n = 200)
ix <- findInterval(df_plot$ndvi, dens$x)
iy <- findInterval(df_plot$lst_delta, dens$y)
df_plot$density <- dens$z[cbind(ix, iy)]

df_plot |>
  ggplot(aes(x = ndvi, y = lst_delta)) +
  geom_point(aes(colour = density), size = 0.4, alpha = 0.6) +
  scale_colour_gradientn(
    colours = c("grey70", "navy", "blue", "cyan", "yellow", "red", "darkred"),
    guide = "none"
  ) +
  geom_smooth(method = "lm", colour = "firebrick") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  coord_cartesian(ylim = c(-10, 10)) +
  labs(title = "NDVI vs ΔLST",
       x = "NDVI", y = "ΔLST [K]") +
  theme_minimal()

cor.test(dry_predictions_lst_ndvi$ndvi, 
         dry_predictions_lst_ndvi$lst_delta,
         method = "spearman",
         use = "complete.obs")

cor.test(dry_predictions_lst_ndvi$ndvi, 
         dry_predictions_lst_ndvi$lst_delta,
         method = "pearson",
         use = "complete.obs")
```

There is a weak negative correlation between dLST and NDVI.

