---
title: "Scatterplots"
author: "Jonathan Lanz"
date: "2026-05-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE, 
                      fig.path = here::here("fig/"), 
                      dev = "png", 
                      dpi = 300
)
```

```{r cars, include=FALSE}
# load libraries and functions
library(ggplot2)
library(here)
library(scico)
library(lubridate)
library(cowplot)
library(dplyr)
source(here::here("R", "get_lst_plot.R"))
source(here::here("R", "get_stats_label.R")) # is executed inside get_lst_plot

# load data
model_data_clean <- readRDS(here::here("data", "model_data_clean.rds")) |>
  filter(!land_cover_type %in% c(11,17))
spatial_predictions_lst <- readRDS(here::here("data", "spatial_predictions_lst.rds"))
dry_predictions_lst <- readRDS(here::here("data", "dry_predictions_lst.rds"))
wet_days <- readRDS(here::here("data", "wet_days.rds"))

```

## Plotting LSTobs vs LSTpot
The following plots show scatterplots of LSTobs vs LSTpot on dry, wet, and all days.
```{r lstobs_lstpot_scatter,  message=FALSE, warning=FALSE, echo=FALSE}
all_predictions_lst <- model_data_clean |>
  filter(!is.na(lst_kelvin)) |>
  dplyr::select(lat, lon, date, pcwd_mm) |>
  right_join(spatial_predictions_lst, by = c("lat", "lon", "date"))

#---plotting LSTobs vs LSTpot---
all_plot_obs <- make_lst_plot(all_predictions_lst,
                          obs_col  = "lst_kelvin",
                          pred_col = "lst_predicted",
                          title_label = "All days",
                          obs_label = "Observed LST [K]",
                          pred_label = "Potential LST [K]")

moist_plot_obs <- make_lst_plot(all_predictions_lst |> filter(date %in% wet_days),
                            obs_col  = "lst_kelvin",
                            pred_col = "lst_predicted",
                            title_label = "\"Moist\" days",
                            obs_label = "Observed LST [K]",
                            pred_label = "Potential LST [K]")

dry_plot_obs <- make_lst_plot(dry_predictions_lst,
                          obs_col = "lst_kelvin",
                          pred_col = "lst_predicted",
                          title_label = "\"Dry\" days",
                          obs_label = "Observed LST [K]",
                          pred_label = "Potential LST [K]")


# combine
combined_lst_comparison_obs <- plot_grid(
  all_plot_obs, moist_plot_obs, dry_plot_obs,
  ncol = 2,
  labels = c("(a)", "(b)", "(c)"),
  label_size = 11,
  label_fontface = "plain",
  rel_widths = c(1, 1, 1))

combined_lst_comparison_obs
```

## Plotting LSTact vs LSTpot
The following scatterplots show the same with LSTact vs LSTpot.

```{r lstact_lstpot_scatter,  message=FALSE, warning=FALSE, echo=FALSE}
#---plotting LSTact vs LSTpot---
all_plot_act <- make_lst_plot(all_predictions_lst,
                              obs_col  = "lst_act",
                              pred_col = "lst_predicted",
                              title_label = "All days",
                              obs_label = "Actual LST [K]",
                              pred_label = "Potential LST [K]")

moist_plot_act <- make_lst_plot(all_predictions_lst |> filter(date %in% wet_days),
                                obs_col  = "lst_act",
                                pred_col = "lst_predicted",
                                title_label = "\"Moist\" days",
                                obs_label = "Actual LST [K]",
                                pred_label = "Potential LST [K]")

dry_plot_act <- make_lst_plot(dry_predictions_lst,
                              obs_col = "lst_act",
                              pred_col = "lst_predicted",
                              title_label = "\"Dry\" days",
                              obs_label = "Actual LST [K]",
                              pred_label = "Potential LST [K]")


# combine
combined_lst_comparison_act <- plot_grid(
  all_plot_act, moist_plot_act, dry_plot_act,
  ncol = 2,
  labels = letters[1:4],
  label_size = 11,
  label_fontface = "plain",
  rel_widths = c(1, 1, 1))

combined_lst_comparison_act
```

## Plotting LSTact vs LSTobs
Finally, the following scatterplots show the same with LSTact vs LSTobs.

```{r lstact_lstobs_scatter, message=FALSE, warning=FALSE, echo=FALSE}
#---plotting LSTact vs LSTobs
all_plot_actvsobs <- make_lst_plot(all_predictions_lst,
                              obs_col  = "lst_act",
                              pred_col = "lst_kelvin",
                              title_label = "All days",
                              obs_label = "Actual LST [K]",
                              pred_label = "Observed LST [K]")

moist_plot_actvsobs <- make_lst_plot(all_predictions_lst |> filter(date %in% wet_days),
                                obs_col  = "lst_act",
                                pred_col = "lst_kelvin",
                                title_label = "\"Moist\" days",
                                obs_label = "Actual LST [K]",
                                pred_label = "Observed LST [K]")

dry_plot_actvsobs <- make_lst_plot(dry_predictions_lst,
                              obs_col = "lst_act",
                              pred_col = "lst_kelvin",
                              title_label = "\"Dry\" days",
                              obs_label = "Actual LST [K]",
                              pred_label = "Observed LST [K]")


# combine
combined_lst_comparison_actvsobs <- plot_grid(
  all_plot_actvsobs, moist_plot_actvsobs, dry_plot_actvsobs,
  ncol = 2,
  labels = letters[1:4],
  label_size = 11,
  label_fontface = "plain",
  rel_widths = c(1, 1, 1))

combined_lst_comparison_actvsobs
```

It is evident that model performance is significantly better on wet days and systematically underpredicts LST on dry days. This is consistent with the expectations, since the model was trained only on wet days.

The most important plots for my analysis are combined in the next plot.

```{r, lstact_lstobs_lstpot_final_scatter}
tight <- theme(plot.margin = margin(2, 2, 2, 2, unit = "pt")) # this was made by Claude

combined_lst_comparison_final <- plot_grid(
  all_plot_obs + tight, 
  moist_plot_obs + tight, 
  dry_plot_obs + tight, 
  all_plot_actvsobs + tight,
  ncol = 2,
  labels = letters[1:4],
  label_size = 11,
  label_fontface = "plain",
  align = "hv",
  axis = "tblr")

combined_lst_comparison_final

```




